Taking Dyads Seriously

### 2018-02-09 paper/1_intro.tex
intro changes

In this article, we present the Additive and Multiplicative Effects (AME) model, a Bayesian approach for directly modeling relational data to better reflect the interdependencies underlying the data generating process of dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still important to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict with a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third order dependencies capture relationships of transitivity, balance, and clusterability between different dyads -- for example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003, if we  understand that the United States invaded Iraq in 2003 and we understand that Poland often participates in US led coalitions. The multiplicative effects capture these sorts of dependencies, especially those that result because the specified model has not accounted for some latent set of shared attributes possessed by actors that affect their probability of interacting with one another.

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, \textit{International Studies Quarterly} published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, so as to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012,franzese:hayes:2007,king:roberts:2014}.

We begin with a discussion of these dependencies and an introduction to the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the context of dyadic data. Last, to highlight the utility of this approach, we apply the AME model to five recent studies in the international relations (IR) literature. The comparison reveals that in accounting for observational dependence AME produces more precise estimates and better calibrated confidence intervals for key variables in the literature. Consequently, AME produces results that, at times, differ from those found in these studies in particular as well as the broader literatures from which they are drawn. Moreover, we demonstrate the latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies in most relational data that the field of IR is concerned with. Finally, we show that for each replication our network-based approach provides substantively more accurate out-of-sample predictions than the models used in the original studies.

The framework that we present advances statistical analysis of dyadic data by accounting for observational dependence while easily testing the substantive effect of variables of interest. Thus, the AME approach can be used by scholars in the field to not only generate substantive insights, but to better model the data generating process behind events of interest. Most importantly, it concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.

>>>>

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, \textit{International Studies Quarterly} published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012,franzese:hayes:2007,king:roberts:2014}.

In this article, we present the Additive and Multiplicative Effects (AME) model, a Bayesian approach for directly modeling relational data to reflect better the interdependencies underlying the data generating process of dyadic data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still essential to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict with a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third order dependencies capture relationships of transitivity, balance, and clusterability between different dyads -- for example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003 if we understand that the United States invaded Iraq in 2003 and that Poland often participates in US-led coalitions. The multiplicative effects capture these sorts of dependencies, especially those that result because the specified model has not accounted for some latent set of shared attributes possessed by actors that affect their probability of interacting with one another.

We begin with a discussion of these dependencies and an introduction to the AME model. Next, we conduct a simulation study to show how the AME approach can recover unbiased and well-calibrated regression coefficients in the context of dyadic data. Last, to highlight the utility of this approach, we apply the AME model to five recent studies in the international relations (IR) literature. The comparison reveals that in accounting for observational dependence, AME produces more precise estimates and better-calibrated confidence intervals for key variables in the literature. Consequently, AME produces results that, at times, differ from those found in these studies in particular as well as the broader literature from which they are drawn. Moreover, we demonstrate the latent factor approach offers substantive insights that are often occluded by ignoring the interdependencies found in the relational data of IR studies. Finally, we show that for each replication our network-based approach provides substantively more accurate out-of-sample predictions than the models used in the original studies.

The framework that we present advances statistical analysis of dyadic data by accounting for observational dependence while easily testing the substantive effect of variables of interest. Thus, the AME approach can be used by scholars in the field to not only generate substantive insights, but to sufficiently model the data generating process behind political events of interest. Most importantly, it concentrates on the relational aspect of the field of international relations through a statistical framework that is familiar to most scholars.
